Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-15 write-ups/PLDI/paper.tex
Finished pass over implementation.

\begin{abstract}
There is growing interest in accelerating irregular data-parallel
algorithms on GPUs.  These algorithms are typically \emph{blocking},
so require fair scheduling.  But GPU programming models (e.g.\ OpenCL)
do not mandate fair scheduling, and GPU schedulers are inherently
unfair in practice.  Current approaches avoid this issue by exploiting
scheduling quirks particular to today's GPUs in a manner that does not
allow the GPU to be shared with other workloads (such as graphics
rendering tasks).  We propose \emph{cooperative kernels}, an extension
to the traditional GPU programming model geared towards writing
blocking algorithms.  Workgroups of a cooperative kernel \emph{are}
fairly scheduled, and multitasking is supported via a small set of
language extensions that the kernel and scheduler use to cooperate.
We describe the semantics of the cooperative kernels programming model
and our prototype implementation.  We evaluate the approach by porting
a set of irregular work stealing and graph algorithms to our
programming model and examining their performance under multitasking
across three GPUs.  Our implementation exploits no
vendor-specific hardware, driver or compiler support, so our encouraging
results provide a lower-bound on the efficiency with
which cooperative kernels can be implemented in practice.

@@

\paragraph{A degree of fairness: occupancy-bound execution}
GPU schedulers are \emph{not} fair. The current GPU
programming models---OpenCL~\cite{opencl2Spec}, CUDA~\cite{cuda-75}  and
HSA~\cite{HSAprogramming11}---specify almost no guarantees regarding scheduling of
workgroups, and current GPUs do not
provide fair scheduling in practice.  Roughly speaking, each workgroup
executing a GPU kernel is mapped to a hardware \emph{compute
  unit}.\footnote{In practice, depending on the kernel, multiple workgroups might map to the same compute unit; we ignore this in our current discussion.}
%
The simplest way for a GPU driver to handle 
the case where more workgroups are launched than there are compute
units is via an \emph{occupancy-bound}
execution model~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16}.  With this model, once a workgroup has commenced execution on a
compute unit (it has become \emph{occupant}), the workgroup has
exclusive access to the compute unit until it has finished execution.
Experiments suggest that the occupancy-bound execution model is widely
implemented by today's GPU devices and drivers~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/oopsla/PaiP16,BNP12}.   It is
evidently a simple model for device vendors to implement efficiently.

@@

\paragraph{Resistance to guaranteeing occupancy-bound execution}

@@

\paragraph{Our proposal: cooperative kernels}
%
To summarise: blocking algorithms
demand fair scheduling of workgroups, but for good reasons
GPU vendors will not commit even to the guarantees of the
occupancy-bound execution model.

@@

In summary, our main contributions are:

\vspace{-1mm}
\begin{itemize}

\item \emph{Cooperative kernels}, an extended GPU programming model that support the scheduling requirements of blocking algorithms (\mysec\ref{sec:cooperativekernels}).

\item A proof-of-concept portable implementation of cooperative
  kernels on top OpenCL 2.0
  (\mysec\ref{sec:implementation}).

\item Experimental results assessing the overhead and responsiveness of the cooperative kernels approach over a set of irregular algorithms across three GPUs (\mysec\ref{sec:experiments}).

\end{itemize}
\vspace{-1mm}

@@

In \mysec\ref{sec:semantics}, we defined the semantics of a resizing barrier in terms
of a series of calls to $\globalbarrier$, $\offerfork$ and $\offerkill$.
 It is possible, however to implement the resizing with
only one call to a global barrier with $\offerfork$ and $\offerkill$
baked into the implementation. \TSComment{note here that these
  implementations require no cfork or ckill outside of the barrier}

We consider barriers which have the master/slave model
(e.g.\ \cite{XF10}). That is, one workgroup (the master) collects signals
from the other workgroups (the slaves) indicating that they have
arrived at the barrier and are waiting for a reply indicating that
they may leave the barrier. Once the master has recieved a signal from
all the slaves, it replies with a signal saying that the slaves may
leave. \TSComment{Need to discuss new active workgroup number that
is stable across iterations}

A first-attempt at incorporating $\offerfork$ and $\offerkill$ into
such a barrier is straight forward. Upon entering the barrier, the
slaves first execute $\offerkill$, possibly exiting. The master then
waits for $M$ slaves (the number of active workgroups), which may
decrease, (based on the $\offerkill$ instructions executed by the
slaves), but will not increase. Once the master observes $M$ slaves
have arrived, it knows all other workgroups are in the kernel are
spinning, waiting to be released. The master executes $\offerfork$,
where there is a conditional for the potentially newly spawned threads
to join the other slaves waiting to be released. Finally, the master
releases all the slaves: the original slaves and the new slaves that
joined at $\offerfork$.

While this barrier implementation works, it has its flaws. Workgroups
only execute $\offerkill$ once per global barrier call. Depending on
the order in which workgroups arrive at the barrier, it is possible
that only one workgroup is killed per barrier call which could lead to
a high time for the scheduler to gather the sufficient amount of
workgroups to execute another kernel concurrently.

We observe if we provide an additional scheduler functionality for a
workgroup executing a cooperative kernel, we can greatly reduce the
gather time. We call the new function {\tt query()}. The semantics
simply return the number of workgroups that the scheduler requires to
kill from the cooperative kernel. A new barrier can now be implented
as follows: (1) Master waits for all slaves; (2) Master forks and sets
new workgroups as slaves. (3) Master calls {\tt query()} and gets
value $W$; (4) master releases slaves and broadcasts the value from
{\tt query} to slaves. (5) Workgroups with id greater than $M-W$ spin
on $\offerkill$, as they know from {\tt query()} that the scheduler
needs them to exit. We show in \mysec\ref{sec:responsiveness} that the barrier using
query greatly reduces the gather time in practice.

@@

We have proposed \emph{cooperative kernels}, a small set of extensions
to the GPUP programming model that allow long-running, blocking
kernels to be fairly scheduled and to share GPU resources with other
workloads via multitasking.  Our prototype implementation uses a
megakernel to mock up the environment in which a cooperative kernel
would run if native support for our programming model was provided.
Our experimental results over three Intel GPUs show that the
programming model is a good fit for the needs of current irregular
algorithms that operate on graphs and use work stealing.  The
performance that could be gained through a native implementation with
dedicated driver support would be even better.

There are three main avenues for future work at present: seeking out
additional classes of irregular algorithms that would benefit from GPU
acceleration using cooperartive kernels, investigating providing
native driver support for cooperative kernels in an open source
graphics driver, and integrating cooperative kernels into the back-end
of a compiler that generates irregular OpenCL from a higher-level DSL,
e.g.\ an OpenCL version of a recent compiler for graph algorithms that
targets CUDA~\cite{DBLP:conf/oopsla/PaiP16}.  Longer-term, as more
vendors adopt the advanced features of OpenCL that our approach
requires, we will be in a position to assess the programming model
across a wider range of devices.

>>>>

\vspace{-3mm}
\begin{abstract}
There is growing interest in accelerating irregular data-parallel
algorithms on GPUs.  These algorithms are typically \emph{blocking},
so require fair scheduling.  But GPU programming models (e.g.\ OpenCL)
do not mandate fair scheduling, and GPU schedulers are inherently
unfair in practice.  Current approaches avoid this issue by exploiting
scheduling quirks particular to today's GPUs in a manner that does not
allow the GPU to be shared with other workloads (such as graphics
rendering tasks).  We propose \emph{cooperative kernels}, an extension
to the traditional GPU programming model geared towards writing
blocking algorithms.  Workgroups of a cooperative kernel \emph{are}
fairly scheduled, and multitasking is supported via a small set of
language extensions that the kernel and scheduler use to cooperate.
We describe the semantics of the cooperative kernels programming model
and our prototype implementation.  We evaluate the approach by porting
a set of irregular work stealing and graph algorithms to our
programming model and examining their performance under multitasking
across three GPUs.  Our implementation exploits no
vendor-specific hardware, driver or compiler support, so our encouraging
results provide a lower-bound on the efficiency with
which cooperative kernels can be implemented in practice.

@@

\vspace{-1mm}
\paragraph{A degree of fairness: occupancy-bound execution}
GPU schedulers are \emph{not} fair. The current GPU
programming models---OpenCL~\cite{opencl2Spec}, CUDA~\cite{cuda-75}  and
HSA~\cite{HSAprogramming11}---specify almost no guarantees regarding scheduling of
workgroups, and current GPUs do not
provide fair scheduling in practice.  Roughly speaking, each workgroup
executing a GPU kernel is mapped to a hardware \emph{compute
  unit}.\footnote{In practice, depending on the kernel, multiple workgroups might map to the same compute unit; we ignore this in our current discussion.}
%
The simplest way for a GPU driver to handle 
the case where more workgroups are launched than there are compute
units is via an \emph{occupancy-bound}
execution model~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16}.  With this model, once a workgroup has commenced execution on a
compute unit (it has become \emph{occupant}), the workgroup has
exclusive access to the compute unit until it has finished execution.
Experiments suggest that the occupancy-bound execution model is widely
implemented by today's GPU devices and drivers~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/oopsla/PaiP16,BNP12}.   It is
evidently a simple model for device vendors to implement efficiently.

@@

\vspace{-1mm}
\paragraph{Resistance to guaranteeing occupancy-bound execution}

@@

\vspace{-1mm}
\paragraph{Our proposal: cooperative kernels}
%
To summarise: blocking algorithms
demand fair scheduling of workgroups, but for good reasons
GPU vendors will not commit even to the guarantees of the
occupancy-bound execution model.

@@

In summary, our main contributions are:
\emph{cooperative kernels}, an extended GPU programming model that supports the scheduling requirements of blocking algorithms (\mysec\ref{sec:cooperativekernels}); a \emph{proof-of-concept implementation} of cooperative
  kernels on top OpenCL 2.0
  (\mysec\ref{sec:implementation}); \emph{experimental results} assessing the overhead and responsiveness of the cooperative kernels approach over a set of irregular algorithms across three GPUs (\mysec\ref{sec:experiments}).

@@

In \mysec\ref{sec:semantics}, we defined the semantics of a resizing barrier in terms
of a series of calls to $\globalbarrier$, $\offerfork$ and $\offerkill$.
 It is possible, however to implement the resizing with
only one call to a global barrier with $\offerfork$ and $\offerkill$
baked into the implementation. \TSComment{note here that these
  implementations require no cfork or ckill outside of the barrier} \ADComment{Ignore until camera-ready?}

We consider barriers that use the master/slave model
(e.g.\ \cite{XF10}): one workgroup (the master) collects signals
from the other workgroups (the slaves) indicating that they have
arrived at the barrier and are waiting for a reply indicating that
they may leave the barrier. Once the master has recieved a signal from
all the slaves, it replies with a signal saying that the slaves may
leave. \TSComment{Need to discuss new active workgroup number that
is stable across iterations} \ADComment{Again, ignore until camera-ready?}

A first-attempt at incorporating $\offerfork$ and $\offerkill$ into
such a barrier is straight forward. Upon entering the barrier, the
slaves first execute $\offerkill$, possibly exiting. The master then
waits for $M$ slaves (the number of active workgroups), which may
decrease due to $\offerkill$ calls by the
slaves, but will not increase. Once the master observes that $M$ slaves
have arrived, it knows that all other workgroups are spinning, waiting to be released. The master executes $\offerfork$,
and the statement immediately following this $\offerfork$ is a conditional that forces newly spawned workgroups to join the
slaves in waiting to be released. Finally, the master
releases all the slaves: the original slaves and the new slaves that
joined at $\offerfork$.

This barrier implementation is suboptimal becuase workgroups only
execute $\offerkill$ once per global barrier call and, depending on
the order in which workgroups arrive at the barrier, it is possible
that only one workgroup is killed per barrier call.  This may not
allow the scheduler to gather workgroups required to execute another
task in a timely fashion.

We can greatly reduce the gather time by providing a new $\keyword{query}()$
function for a workgroup executing a cooperative kernel, which returns the number of workgroups that the scheduler needs to obtained
from the cooperative kernel.  
%
The programmer cannot call $\keyword{query}$ directly;
it is only invoked from within the resizing barrier implementation.
%
A resizing barrier can now be implemented
as follows: (1) the master waits for all slaves to arrive; (2) the master calls $\offerfork$ and
commands the new workgroups to be slaves; (3) the master calls $\keyword{query}$, obtaining a value $W$;
(4) the master releases the slaves, broadcasting the value $W$ to them; (5) workgroups with ids larger than $M-W$ spin, calling 
$\offerkill$ repeatedly until the scheduler claims them---we know from $\keyword{query}$ that the scheduler will eventually do so.
We show in \mysec\ref{sec:responsiveness} that the barrier using
query greatly reduces the gather time in practice.

@@

We have proposed \emph{cooperative kernels}, a small set of GPU
programming extensions that allow long-running, blocking kernels to be
fairly scheduled and to share GPU resources with other workloads.
Experimental results using our megakernel-based prototype over three
Intel GPUs shows that the model is a good fit for the needs of current
GPU-accelerated irregular algorithms.  The
performance that could be gained through a native implementation with
dedicated driver support would be even better.

There are three main avenues for future work at present: seeking
additional classes of irregular algorithms to which the model might
(be extended to) apply (to), investigating native driver support for
cooperative kernels via open source graphics drivers, and integrating
cooperative kernels into template- and compiler-based programming
models for graph algorithms on
GPUs~\cite{SreesCompetition,DBLP:conf/oopsla/PaiP16}.  As more vendors
adopt the advanced features of OpenCL that our approach requires, we
will be in a position to assess the programming model across a wider
range of devices.
